Achieving maximum possible precision on stellar radial-velocity measurements in multi-epoch spectroscopy

### 2017-11-28 paper/paper.tex
outline

\begin{abstract}
% context
Extremely precise stellar radial velocity (\RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
Despite the importance of these observations, the detailed computation methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
% aims
In this work, we look at sensible approaches to extracting \RV s from a data set of  one-dimensional, calibrated, extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
% methods
We consider cross-correlations (and, equivalently, maximum-likelihood optimizations) of the data with generalized (weighted) binary masks, synthetic spectra (from, say, stellar models), and data-driven templates constructed from the data themselves.
We also consider the possibility that there are low-amplitude, unknown telluric absorptions (micro-tellurics) affecting the spectrum, and continuum-normalization problems, and measure improvements that flow from modeling those also in a data-driven way.
% results
We find that of all methods, only cross-correlation with an iteratively built data-driven template comes close to saturating the \CRLB.
We find that if there are micro-tellurics, they can be captured by a data-driven model.
We predict that simultaneously fitting data-driven models of the stellar spectrum and telluric contributions at all observational epochs will improve the \RV\ precision of \HARPS\ to substantially better than $1\,\ms$.
\end{abstract}

@@

@@

cross-correlation vs. max likelihood

for either of these, you could use:
\begin{itemize}
\item ``binary'' mask
\item synthetic spectrum (or template observation?)
\item data-driven template spectrum
\end{itemize}

information theory based discussion of these: compute the CRLB for velocity. Mention how uncertainties are typically computed (Butler 1996 and Bouchy 2001 formulae).

@@

\section{Testing with Toy-Model Spectra}

@@

We repeated this synthesis to make a set of 512 spectra, each with a random \RV\ between $-30$ and $30\,\kms$ (the approximate amplitude of the yearly \RV\ shifts induced by barycentric motion). The \RV s can be applied as a straightforward Doppler shift in the centers of each line. \todo{(add equation?)} At this point, we are able to run a cross-correlation or a maximum-likelihood analysis on the set of spectra and examine the deviations in the recovered \RV s from the input \RV s.

\todo{remember to discuss quadratic max procedure.. probably in previous section?}

\subsection{Single-Line Results}

@@

\subsection{Multi-Line Results}

@@

\section{Dependence of RV Precision on Template Choice}

how well does the mask/template approximate the real data? try fudging the relative line depths and locations in both the binary mask and the template.

is this where we bring in the data-driven template and demonstrate how well it performs?

\section{Correlated Noise Sources and Their Consequences}

additional noise considerations:
\begin{itemize}
\item (micro-) telluric features
\item subtle continuum trends
\item cosmic rays?
\end{itemize}

\section{Application to Real Spectra}

how do our results compare to the reality of the situation? (will this tell us anything other than ``real spectra are more complicated and our template is silly''?)

%\section{Discussion}

\section{Conclusions}

\acknowledgements
MPIA hospitality

>>>>

\begin{abstract}
% context
Extremely precise stellar radial velocity (\RV) measurements are a foundational tool in the field of exoplanets, with hundreds of planets discovered by the Doppler method to date and about a dozen high-resolution spectrographs around the world dedicated to the search.
Despite the importance of these observations, the detailed computational methods used to extract $\ms$-level \RV s from stellar spectra remain largely unstandardized and unpublished.
% aims
In this work, we look at sensible approaches to extracting \RV s from a data set of extracted spectra and compare their efficiencies by testing on a segment of realistic model spectrum, with the goal of identifying methods that have the possibility of saturating the information-theoretic Cram\'er--Rao bound (\CRLB) on measurement precision.
% methods
We consider cross-correlations (and, equivalently, maximum-likelihood optimizations) of the data with generalized (weighted) binary masks, synthetic spectra (from, say, stellar models), and data-driven templates constructed from the data themselves.
We also consider the possibility that there are low-amplitude, unknown telluric absorptions (micro-tellurics) affecting the spectrum, and continuum-normalization problems, and measure improvements that flow from modeling those also in a data-driven way.
% results
We find that of all methods, only cross-correlation with an iteratively built data-driven template comes close to saturating the \CRLB.
We find that if there are micro-tellurics, they can be captured by a data-driven model.
We predict that simultaneously fitting data-driven models of the stellar spectrum and telluric contributions at all observational epochs will improve the \RV\ precision of \HARPS\ to substantially better than $1\,\ms$.
\end{abstract}

@@

The assumption of this project is that we have many observations of the same star with good coverage throughout the observing season, so that we have seen the star many times spanning the range of telluric-to-star relative RV shifts. This means we are not considering the case of a real-time data reduction pipeline. We also assume a HARPS-like setup, i.e. no absorption cell, but the points we make can be easily generalized to the gas-cell case. We return to this point in the discussion \todo{(do this)}.

We additionally assume that the data are extremely well wavelength-calibrated. In other work, we will discuss this, but our current belief is that current RV spectrographs are supremely well wavelength-calibrated.

\section{Information Theory}
\label{s:info}

\begin{itemize}
\item the Cramer-Rao bound is the limit on measurement
\item it relates to Fisher information
\item it cannot be computed without a model
\item CRLB is saturated with a maximum likelihood estimator under the correct model
\item wrong spectrum template = wrong and noisy \RV s
\end{itemize}

\begin{itemize}
\item what is the Cramer-Rao bound on a spectrum?
\item Butler equation
\item Bouchy equation
\item Figueira (or Lovis \& Fischer) equation
\end{itemize}

In conclusion: the saturate the CRLB, we need to construct a very accurate spectral template, an accurate tellurics template, and maximize the likelihood. Given that theory sucks, this probably means data-driven.

@@

cross-correlation and max likelihood are the same: here's why. We are going to consider cross-correlation for simplicity in this paper.

for either of these, you could use:
\begin{itemize}
\item ``binary'' mask: reasons you might want to use this (automatically ignores tellurics, trivially computes average line profile), pipelines that use this \& how (quote HARPS papers)
\item synthetic spectrum (i.e. a slightly wrong model prediction, since the stellar model will never be a perfect reproduction of the star in question): sub-tweet: some surveys cross-correlate with a single (same) template for every single star, which is extremely wrong. you can be more right by using a spectral-type-specific template.
\item data-driven template spectrum: some surveys (HIRES) use a single observation as a template. combining information from many visits is better because the combined SNR is amazing.
\end{itemize}

discuss quadratic max procedure to get a fine-grained maximum from a coarse-grained cross-correlation grid

@@

\section{Data}

Most of what we do in this paper uses artificial data because we need access to the truth in order to compute the theoretical information content of the data. As a final test, we do use real data as a demonstration that the methods presented here are useful when applied to real data and the assumptions made in our artificial data generation are realistic.

\subsection{Artificial Data}

@@

We repeated this synthesis to make a set of \todo{64} spectra, each with a random \RV\ between $-30$ and $30\,\kms$ (the approximate amplitude of the yearly \RV\ shifts induced by barycentric motion). The \RV s can be applied as a straightforward Doppler shift in the centers of each line. \todo{(add equation?)} At this point, we are able to run a cross-correlation or a maximum-likelihood analysis on the set of spectra and examine the deviations in the recovered \RV s from the input \RV s.

discuss single-line and multi-line cases

FIGURE showing multi-line artificial data

\subsection{HARPS Data}

munging notes

FIGURE showing analogous section of real data

\section{Experiments}

\subsection{Single-Line}

@@

\subsection{Binary Mask}

@@

\subsection{Rigid Template}

how well does the mask/template approximate the real data? try fudging the relative line depths and locations in both the binary mask and the template. do a correct template and an incorrect template.

even a tiny wrongness to the template is visible in the RV variance.

\subsection{Data-Driven Template}

how do we construct the template? show that it comes close to saturating the CRLB.

comment: there are far better ways to derive the data-driven template. we leave this to future work.

\subsection{Realistic Sources of Structured Noise}

additional noise considerations:
\begin{itemize}
\item (micro-) telluric features: tellurics that you can't identify are important
\item subtle continuum trends: not that big a deal
\item cosmic rays?: we might skip this.
\end{itemize}

\subsection{HARPS Data}

all the data-driven code works on real data. we get a better RMS on the star from the data-driven template than from the binary mask. based on this, we believe that the HARPS pipeline doesn't strictly use a binary mask.

\section{Summary \& Discussion}

paragraph for each major result from the abstract

wrongness of our assumptions - how does this affect our conclusions?

\acknowledgements
MPIA hospitality, Montet, Foreman-Mackey, Bean, Sturmer, Rix, HARPS team, ESO, Stars group meeting, Brewer
